---
title: "Modelling a Single-Trial ERP Waveform"
subtitle: "Basis-function regression, and the choice of noise model"
author: "Mark Andrews"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(nlme)
knitr::opts_chunk$set(fig.width = 8, fig.height = 4.5, fig.align = "center")
accent <- "#BE0000"
```

# The problem

Take the simplest possible version of the modelling problem. One
subject, one trial, one electrode. The data are a single voltage
waveform $y(t)$ observed at $T$ timepoints spanning $-200$ to $1000$ ms
relative to stimulus onset. There is no averaging anywhere. This is
what a single row of the eventual multilevel model has to deal with,
and if the noise model is wrong here it will be wrong everywhere above
it.

The waveform is a nonlinear function of time. We want to describe it
as a smooth function plus something else, and the whole substance of
this document is what the something else should be.

# The mean model

## Basis-function regression

A nonlinear function of time can be written as a linear combination of
$K$ fixed nonlinear functions of time,

$$
f(t) \;=\; w_0 \;+\; \sum_{k=1}^{K} w_k \, \phi_k(t).
$$

The $\phi_k$ are deterministic and fixed in advance. The $w_k$ are
ordinary regression coefficients. The model is nonlinear in $t$ but
linear in $w$, so all the machinery of linear regression applies
unchanged. Polynomial regression, spline regression, and Fourier
regression are all instances of this. They differ only in the choice
of $\phi_k$.

We use Gaussian radial basis functions,

$$
\phi_k(t) \;=\; \exp\!\left( -\frac{(t - \mu_k)^2}{2 \ell_k^2} \right),
$$

with centres $\mu_k$ and widths $\ell_k$ fixed as data rather than
estimated. This keeps the model linear in its parameters.

The choice is deliberate and it is not a spline. The reason is
interpretability. Each $\phi_k$ is a localised bump with a latency
$\mu_k$ and a width $\ell_k$, which are the same quantities in which
ERP components are described. A basis centred at 240 ms with a width
of 40 ms is a P2p-shaped object before any data are seen. A B-spline
basis function has no such reading, and neither does a polynomial
term. Since the scientific questions here are about specific
components at specific latencies, a basis whose elements are already
component-shaped is worth having.

Two densities of basis are used: a coarse set spanning the whole
epoch, and a denser, narrower set concentrated over $0$ to $300$ ms
where P1, N1, and P2p all sit.

```{r}
#| label: basis

rbf_design_matrix <- function(x, centers, width) {
  mapply(function(ctr, w) exp(-0.5 * (x - ctr)^2 / w^2), centers, width)
}
```

## One caveat on conditioning

Overlapping bumps are collinear. With widths equal to the centre
spacing, the design matrix for the 16-function basis used here has a
condition number of about 470, and the fitted weights come out in the
hundreds against data with a standard deviation of 8 µV. There is
heavy cancellation between neighbouring basis functions.

This does not damage prediction, and it does not damage inference
about the fitted curve, but it does mean that individual $w_k$ are not
separately interpretable and that their standard errors are unstable.
Any quantity you want to report should be a function of the fitted
curve, such as $f(240)$ or the height of the fitted P2p peak, rather
than a single weight. Narrowing the widths relative to the spacing
reduces the overlap, and is worth an experiment.

# The noise model

## Why the question matters

Write the model in vector form. Let $y$ be the $T$-vector of voltages,
$B$ the $T \times K$ matrix with $B_{tk} = \phi_k(t)$, and $\epsilon$
the $T$-vector of residuals:

$$
y \;=\; \alpha \mathbf{1} + B w + \epsilon.
$$

Whatever we assume about $\epsilon$, the least-squares estimate of $w$
remains unbiased. What changes is its variance. Under
$\epsilon \sim \mathrm{MVN}(0, \Sigma)$ the generalised least squares
estimator has

$$
\mathrm{Var}(\hat{w}) \;=\; \left( B^{\top} \Sigma^{-1} B \right)^{-1},
$$

so the noise model enters inference only through $\Sigma^{-1}$. Get
$\Sigma$ wrong and the point estimates are roughly fine while the
uncertainty is wrong, possibly by a large factor. Everything below is
therefore about calibration, not about fit.

## Why iid is not defensible here

The default assumption is
$\epsilon_t \stackrel{\text{iid}}{\sim} N(0, \sigma^2)$, so
$\Sigma = \sigma^2 I$ and

$$
\mathrm{Var}(\hat{w}) \;=\; \sigma^2 (B^{\top} B)^{-1},
$$

which shrinks like $1/T$. The assumption says that knowing
$\epsilon_t$ tells you nothing about $\epsilon_{t+1}$.

This is false, and not marginally so. A single-trial EEG waveform is
visibly smooth. It has to be: the data were band-pass filtered at
$1$-$30$ Hz, so by construction nothing in the record can change faster
than 30 times a second, while the sampling rate is 1024 Hz. On top of
that, the residual after removing a smooth evoked component is
dominated by ongoing EEG, which is itself a strongly autocorrelated
oscillatory process.

Empirically, on the trial analysed here, the residual autocorrelation
under an iid model is 0.77 at lag 1 and 0.34 at lag 2. It then crosses
zero and reaches $-0.46$ by lag 6. The iid model is describing a
process with none of this structure.

## The general form: a multivariate normal residual

The natural generalisation is to treat the residual not as $T$
independent scalars but as a single draw from a $T$-dimensional
Gaussian,

$$
\epsilon \;\sim\; \mathrm{MVN}\!\left(0, \; \sigma^2 R \right),
\qquad R_{tt'} \;=\; \kappa(t, t'),
$$

where $\kappa$ is a correlation function. If the process is stationary,
$\kappa$ depends only on the lag, $\kappa(t, t') = \rho(|t - t'|)$, and
$R$ is Toeplitz. Every model below is a choice of $\rho$.

## Why not a Gaussian (RBF) kernel

The obvious first guess is the squared exponential,

$$
\rho(h) \;=\; \exp\!\left( -\frac{h^2}{2\lambda^2} \right).
$$

There are three reasons not to use it here, of increasing severity.

**It has the wrong short-lag behaviour.** Its spectral density is
$\propto \exp(-\lambda^2 \omega^2 / 2)$, decaying faster than any
polynomial. Sample paths are infinitely differentiable. Real EEG, even
after low-pass filtering, is rougher than this, and a process with
Gaussian spectral decay cannot represent that.

**It cannot represent an oscillation.** The squared exponential has a
monotone decreasing spectral density with no peak anywhere, and a
monotone decreasing autocorrelation function that never goes negative.
The observed residual autocorrelation reaches $-0.46$ at lag 6. No
choice of $\lambda$ can produce that. The mismatch is structural, not a
matter of tuning. Since ongoing posterior EEG is dominated by the alpha
rhythm, which is exactly a spectral peak, this rules the kernel out on
physiological grounds as well as empirical ones.

**It is the same object as the mean model.** This is the deepest
objection. Radial basis function regression with $K$ Gaussian bumps and
iid Gaussian weights converges, as $K \to \infty$, to a Gaussian
process with a squared exponential kernel [@neal1996]. So a model with
Gaussian RBFs in the mean and a Gaussian kernel in the residual is a
model in which the residual term is the mean term with $K$ taken to
infinity. The two components are not merely similar. They are the same
model class, competing to explain the same structure, and the split
between them is determined by the priors rather than by the data.

Choosing a residual kernel from a family that cannot be written as a
limit of the mean model is what makes the decomposition mean anything.

## Autoregressive processes

An AR($p$) process is defined by the recursion

$$
\epsilon_t \;=\; \sum_{j=1}^{p} \varphi_j \, \epsilon_{t-j} \;+\; a_t,
\qquad a_t \stackrel{\text{iid}}{\sim} N(0, \sigma_a^2).
$$

There is no intercept, since $\epsilon$ is a mean-zero error process by
construction. The innovation $a_t$ is essential and is the only iid
term anywhere in the model. Everything correlated in $\epsilon$ is the
accumulated history of past shocks passed through the recursion.

The process is $p$th-order Markov: $\epsilon_t$ depends on the past
only through $(\epsilon_{t-1}, \ldots, \epsilon_{t-p})$.

This is not a different kind of model from the multivariate normal
above. It is the same model, written recursively. The autocorrelation
function of a stationary AR($p$) satisfies the Yule-Walker recursion

$$
\rho_h \;=\; \sum_{j=1}^{p} \varphi_j \, \rho_{h-j}, \qquad h \geq 1,
$$

and $R_{tt'} = \rho_{|t - t'|}$ gives the Toeplitz correlation matrix
of the multivariate normal. The recursion and the covariance matrix are
two views of one object.

For the trial analysed here the fitted AR(2) coefficients are
$\varphi_1 = 1.365$, $\varphi_2 = -0.631$, giving

```
       [,1]   [,2]   [,3]   [,4]   [,5]   [,6]   [,7]   [,8]
[1,]  1.000  0.837  0.511  0.170 -0.091 -0.231 -0.258 -0.206
[2,]  0.837  1.000  0.837  0.511  0.170 -0.091 -0.231 -0.258
[3,]  0.511  0.837  1.000  0.837  0.511  0.170 -0.091 -0.231
[4,]  0.170  0.511  0.837  1.000  0.837  0.511  0.170 -0.091
```

Toeplitz, as it must be, but with correlations that fall, cross zero,
and go negative. This is the structure a Gaussian kernel cannot
produce.

## What AR(2) means physically

When the characteristic polynomial $1 - \varphi_1 z - \varphi_2 z^2$
has complex roots, which happens when $\varphi_1^2 + 4\varphi_2 < 0$,
the autocorrelation function is a damped cosine,

$$
\rho_h \;\propto\; \varrho^{\,h} \cos(\omega h + \theta),
\qquad
\varrho = \sqrt{-1/\varphi_2}^{\,-1},
\qquad
\omega = \arccos\!\left( \frac{\varphi_1}{2\sqrt{-\varphi_2}} \right).
$$

This is the discrete-time counterpart of a damped stochastic harmonic
oscillator: a resonant frequency with a Lorentzian spectral peak whose
width is set by the damping. It is the simplest object that can
represent a rhythm.

For the fitted values above, the envelope decays by a factor of 0.794
per sample and $\omega$ corresponds to **10.95 Hz** at a 128 Hz
sampling rate. That is the alpha rhythm. It is worth being clear about
what this is and is not. It is not a fitted target. The model was given
two free parameters and independently recovered the dominant rhythm of
posterior EEG, at a parieto-occipital electrode, where it is exactly
what one would expect to find. It is reasonable evidence that the AR(2)
residual is describing something physiological rather than absorbing
misfit.

Note also that AR(1) cannot do this. Its autocorrelation function is
$\rho_h = \varphi^h$, monotone and never negative. Fitted to this
trial, $\varphi$ runs to $0.9999997$, hard against the boundary, and
the model degenerates into a random walk with a standard error on the
intercept of 5052 against a point estimate of 279. It fails not because
autocorrelation is the wrong idea but because a monotone
autocorrelation function is the wrong shape.

# Why AR($p$), out of infinitely many possibilities

This is the substantive theoretical question, and it deserves a
substantive answer rather than a list of models that happened to fit
worse.

The complaint is fair as stated. AR(2) beating iid, AR(1), and a
squared exponential kernel is four points sampled from an infinite
space. If the argument were only "we tried some and this one won", it
would carry very little weight.

The argument is not that. It is that AR($p$) is not a point in the
space of covariance models. It is a **sieve**: a nested sequence of
model classes whose union is dense in the class of stationary Gaussian
processes.

Two standard results make this precise.

**Wold decomposition.** Any purely non-deterministic weakly stationary
process can be written as an infinite moving average of uncorrelated
innovations, $\epsilon_t = \sum_{j \geq 0} \psi_j a_{t-j}$. If the
corresponding transfer function is invertible, this can be inverted to
an infinite autoregression, $\epsilon_t = \sum_{j \geq 1} \pi_j
\epsilon_{t-j} + a_t$.

**Spectral approximation.** The spectral density of an AR($p$) process
is

$$
S(\omega) \;=\; \frac{\sigma_a^2}{2\pi} \left| 1 - \sum_{j=1}^{p} \varphi_j e^{-\mathrm{i} j \omega} \right|^{-2},
$$

which is a rational function of $e^{-\mathrm{i}\omega}$. Any continuous
spectral density bounded away from zero can be approximated uniformly
by such a rational function, with the approximation error going to zero
as $p$ increases.

The consequence is that choosing AR($p$) is not choosing one kernel out
of infinitely many. It is choosing a family that can approximate any of
them, and then choosing the order $p$. That converts an unstructured
search over an infinite space into a one-dimensional model-selection
problem over a nested sequence, which is a problem we know how to
handle.

The squared exponential kernel is not a member of any such sequence. It
is a single point, with a fixed spectral shape and no dial to turn if
that shape is wrong.

Two further properties make AR($p$) the practical choice rather than
merely a defensible one.

It has an exact state space representation, so the likelihood is
evaluable in $O(Tp)$ by recursion. A general kernel requires a
$T \times T$ Cholesky at $O(T^3)$. For $T = 154$ this is the difference
between milliseconds and seconds; inside an MCMC loop over thousands of
iterations and, eventually, thousands of trials, it is the difference
between feasible and not.

And its low-order members are physically interpretable. AR(1) is the
discretely sampled Ornstein-Uhlenbeck process, equivalently a Matérn
kernel with $\nu = 1/2$. AR(2) with complex roots is a damped harmonic
oscillator. These are not arbitrary. They are the objects one would
write down from first principles to describe broadband and rhythmic
neural activity respectively.

# How do we know it is good enough?

Three criteria, and an honest account of what they say.

## Criterion 1: does the residual model whiten the residuals?

The definitive test. If $\epsilon \sim \mathrm{MVN}(0, \sigma^2 R)$ is
correct, then the normalised residuals $L^{-1}\hat{\epsilon}$, where
$LL^{\top} = R$, are iid standard normal. For an AR model these are the
one-step prediction errors, obtained directly from the recursion. Test
them for whiteness with the autocorrelation function and a Ljung-Box
test. If they are white, the noise model has extracted all the linear
structure and no further elaboration is warranted.

## Criterion 2: nested model comparison

AR($p$) is nested in AR($p+1$), so likelihood ratio tests and
information criteria apply directly.

## Criterion 3: stability of the reported quantity

The scientifically honest criterion. If the conclusion you report,
say the posterior interval for the fitted amplitude at 240 ms, is
materially the same under AR(2), AR(3), and AR(4), then the exact order
does not matter and you need not resolve it. If it changes, that
sensitivity is itself the finding and should be reported.

## What these criteria actually say on a single trial

They do not settle the question, and it would be misleading to suggest
otherwise.

On this trial at 128 Hz, AIC continues to improve monotonically as $p$
increases, from 988 at $p = 0$ to 191 at $p = 8$, with no sign of
levelling off. The Ljung-Box test on the normalised residuals rejects
whiteness at every order tried. The estimated innovation standard
deviation falls steadily, from 5.32 µV at $p = 0$ to 0.35 µV at
$p = 8$.

The reason is the identifiability problem flagged earlier, and it is
worth stating plainly. On a single trial the evoked response and the
ongoing EEG are both smooth low-frequency processes on the same domain,
with no spectral gap between them. A 16-function basis cannot capture
all the smooth structure in the waveform, and a high-order AR is
perfectly capable of capturing the remainder. So as $p$ grows, the
noise model absorbs signal, the innovation variance collapses, and the
likelihood improves without bound. **Order selection by likelihood on a
single trial is not a well-posed problem.**

Criterion 3 is more reassuring. The standard error of the fitted
waveform at the P2p latency behaves like this:

| $p$ | SE of $\hat f(240)$ |
|-----|---------------------|
| 0 (iid) | 2.07 |
| 1 | 4.49 |
| 2 | 3.05 |
| 3 | 3.80 |
| 4 | 3.01 |
| 5 | 3.09 |
| 6 | 3.66 |
| 7 | 2.60 |

The estimable quantity is far more stable than the individual weights,
whose standard errors swing by a factor of three over the same range.
Two things follow. The iid model understates uncertainty on the fitted
curve by roughly 50%, consistently, and that conclusion does not depend
on the AR order. And within $p = 2$ to $7$ the answer is essentially
flat, so the exact order does not need resolving for this purpose.

What this means practically is that AR(2) should be adopted as a
defensible minimum, chosen on physiological grounds (it is the lowest
order that can represent a rhythm, and it recovers alpha) rather than
by AIC, which cannot arbitrate here. Two further steps are then worth
taking.

The first is to estimate the noise parameters from the pre-stimulus
interval, where there is no evoked signal by construction, pooling
across trials to get enough data, and then to fix them or use them as a
tight prior when fitting the full epoch. This separates the two
estimation problems cleanly.

The second, and the real resolution, is replication. In the full
multilevel model the evoked component is phase-locked across trials and
the residual is not. That asymmetry is what identifies the
decomposition. The single-trial ambiguity described here is a property
of the single-trial problem, not of the eventual model.

# Downsampling

## What is being claimed

The recordings are at 1024 Hz, band-pass filtered at 1 to 30 Hz, over
epochs of 1.2 s, giving $T = 1230$ samples per trial. The claim is
that these can be reduced to every eighth sample, $T = 154$ at 128 Hz,
with no material loss.

## Why it works

The sampling theorem. A signal with no spectral content above $f_{\max}$
is completely determined by samples taken at any rate above $2 f_{\max}$.
The band-pass filter puts $f_{\max} = 30$ Hz. Decimating by 8 gives a
new sampling rate of 128 Hz and a new Nyquist frequency of 64 Hz,
comfortably above 30 Hz. The discarded samples are not additional
information. They are the retained samples restated.

Ordinarily decimation requires an anti-alias filter first, so that
content above the new Nyquist does not fold down into lower
frequencies. Here that filter has already been applied as part of
preprocessing, so plain decimation is safe. It would not be safe on
unfiltered data.

## Verifying it empirically

Decimate, interpolate back onto the original grid, and compare. On the
trial analysed here:

| decimation | rate | $T$ | RMSE / SD of signal |
|---|---|---|---|
| by 2 | 512 Hz | 615 | 0.005% |
| by 4 | 256 Hz | 308 | 0.026% |
| by 8 | 128 Hz | 154 | 0.850% |
| by 16 | 64 Hz | 77 | 16.7% |

At 128 Hz the reconstruction error is 0.069 µV against a signal
standard deviation of 8.16 µV. Almost all of that error is at the two
epoch edges, where the interpolating spline has data on only one side.
Restricted to the interior of the epoch the error falls to 0.32% of the
signal SD with a maximum absolute error of 0.08 µV. The P1 component
has an amplitude of several µV, so the components survive with two
orders of magnitude of margin.

The breakdown at 64 Hz is reassuring rather than worrying. It is
exactly twice the filter cutoff, so there is no headroom, and cubic
spline interpolation is not sinc interpolation. The failure confirms
the check is capable of detecting real information loss, which is what
makes the clean result at 128 Hz meaningful.

Confirmation from the periodogram of the same trial: 97.4% of the power
lies below 30 Hz, 2.3% between 30 and 64 Hz (the filter's roll-off
skirt), and 0.3% above 64 Hz. There is essentially nothing above the
new Nyquist frequency to lose.

## Why it is not a matter of fewer unknowns

It is worth being precise about the mechanism, because there is a
natural but mistaken way to describe it.

Downsampling does **not** reduce the number of unknowns. The parameters
are $\alpha$, the $K$ basis weights, $\sigma$, and the $p$
autoregressive coefficients. That count is $K + p + 2$ regardless of
$T$. Decimation changes the number of *observations*, not the number of
things being inferred. If anything, reducing $T$ while holding the
parameter count fixed makes the inference problem nominally harder, not
easier.

The correct statement is about information rather than dimension. The
Fisher information for $w$ is

$$
\mathcal{I}(w) \;=\; B^{\top} \Sigma^{-1} B.
$$

Under the iid model $\Sigma^{-1} = \sigma^{-2} I$ and this grows
linearly in $T$, which is precisely the bug: adding redundant samples
appears to add information without limit. Under a correctly specified
correlated $\Sigma$, the inverse downweights the redundancy, and as the
sampling rate increases past the point where the samples become
predictable from their neighbours, $\mathcal{I}(w)$ approaches a finite
limit set by the underlying continuous-time process. Sampling faster
than the bandwidth adds numbers, not evidence.

So downsampling is close to free not because the model has less to do,
but because a correct noise model already knows those samples were
redundant. The two ideas are the same idea seen from two sides.

## Why it is necessary, not merely permitted

The stronger claim is that decimation is required for the noise model
to be estimable at all, and this is demonstrable.

Fitting AR($p$) to the same trial at the full 1024 Hz gives an
estimated innovation standard deviation of 0.053 µV at $p = 2$ and
approximately 0.001 µV at $p = 8$, with AIC falling to $-16600$. The
model is claiming that the waveform is very nearly deterministic. In a
sense it is right, because a 30 Hz band-limited signal sampled at 1024
Hz genuinely is almost perfectly predictable one step ahead. But the
consequence is that the likelihood becomes dominated by predicting the
deterministic interpolation between samples rather than by anything
about the signal, the autoregressive roots crowd onto the unit circle
($\varphi = (1.986, -0.993)$ at $p = 2$), and the posterior geometry
becomes badly conditioned. Model comparison across orders becomes
meaningless, and any resulting standard errors are not credible.

This is the oversampling degeneracy, and it is the real argument for
decimation. At 1024 Hz you cannot fit a low-order autoregressive noise
model without pathology. At 128 Hz you can.

## The computational consequence

The likelihood is $O(Tp)$, so decimating by 8 reduces the cost of a
single gradient evaluation by a factor of 8. In the full multilevel
model the total cost scales as
$N_{\text{subjects}} \times N_{\text{trials}} \times T$, so the factor
of 8 applies to the whole thing.

The realised speedup will be at least 8 and probably more, because the
number of gradient evaluations Hamiltonian Monte Carlo needs depends on
the posterior geometry, and the geometry at 1024 Hz is far worse for
the reason just described. But the honest framing is not that
decimation makes a feasible model 8 times faster. It is that it makes
an otherwise degenerate model well behaved, and the speedup comes on
top.

# Implementation in Stan

## Parameterising the AR coefficients

Two things need care.

**Stationarity.** The AR coefficients must lie in the stationarity
region, the set for which the roots of the characteristic polynomial
lie outside the unit circle. For $p = 2$ this is a triangle; for
general $p$ it is an awkward region that cannot be expressed as simple
box constraints. Rejection sampling on the constraint produces a
discontinuous posterior and poor Hamiltonian Monte Carlo behaviour.

The solution is to parameterise by the **partial autocorrelations**
$r_1, \ldots, r_p$, each free in $(-1, 1)$, and map them to the AR
coefficients with the Levinson-Durbin recursion. The image of the open
cube $(-1,1)^p$ under this map is exactly the stationarity region
[@bnschou1973]. Stationarity is then enforced by construction, with a
smooth unconstrained geometry.

**Exact rather than conditional likelihood.** The easy option is to
condition on the first $p$ observations. The exact likelihood requires
the stationary distribution of the first $p$, which in general means a
$p \times p$ solve.

The same Levinson-Durbin recursion supplies this for free. At time $t$,
the best linear predictor of $\epsilon_t$ from its $t-1$ predecessors
uses the order-$\min(t-1, p)$ coefficients that the recursion produces
along the way, with prediction variance
$\sigma^2 \prod_{i \leq \min(t-1,p)} (1 - r_i^2)$. Accumulating these
one-step prediction errors gives the exact log-likelihood, with no
matrix ever formed and no Cholesky. This decomposition was verified
against an explicit multivariate normal density and agrees to machine
precision.

Note that $\sigma$ in this parameterisation is the **marginal**
standard deviation of $\epsilon$, not the innovation standard
deviation. The two differ by $\prod_i (1 - r_i^2)^{1/2}$. The marginal
scale is the more interpretable of the two and the easier to put a
prior on, since it is on the scale of the observed data.

Setting $p = 0$ recovers iid noise exactly, so one file fits both
models.

## The model

```{.stan filename="stan/rbf_arp.stan"}
data {
  int<lower=1> T;                    // number of timepoints
  int<lower=1> K;                    // number of basis functions
  int<lower=0> p;                    // AR order; p = 0 gives iid noise
  matrix[T, K] B;                    // RBF design matrix (no intercept column)
  vector[T] y;                       // observed voltage
  real<lower=0> prior_sd_w;          // prior sd for basis weights
  real<lower=0> prior_sd_sigma;      // prior scale for residual sd
}

parameters {
  real alpha;                        // intercept
  vector[K] w;                       // basis weights
  real<lower=0> sigma;               // MARGINAL sd of eps
  vector<lower=-1, upper=1>[p] r;    // partial autocorrelations
}

transformed parameters {
  // Levinson-Durbin: partial autocorrelations -> AR coefficients at
  // every order up to p. a[k, 1:k] holds the order-k coefficients.
  matrix[p, p] a = rep_matrix(0, p, p);
  vector[p + 1] v;                   // v[k+1] = prod_{i<=k} (1 - r[i]^2)

  v[1] = 1;
  for (k in 1 : p) {
    a[k, k] = r[k];
    for (j in 1 : (k - 1)) {
      a[k, j] = a[k - 1, j] - r[k] * a[k - 1, k - j];
    }
    v[k + 1] = v[k] * (1 - square(r[k]));
  }
}

model {
  vector[T] eps = y - alpha - B * w;

  alpha ~ normal(0, prior_sd_w);
  w ~ normal(0, prior_sd_w);
  sigma ~ normal(0, prior_sd_sigma);         // half-normal via <lower=0>
  r ~ uniform(-1, 1);                        // flat on the stationarity region

  // Exact likelihood by prediction-error decomposition.
  for (t in 1 : T) {
    int k = min(t - 1, p);                   // order usable at this t
    real mu = 0;
    for (j in 1 : k) {
      mu += a[k, j] * eps[t - j];
    }
    target += normal_lpdf(eps[t] | mu, sigma * sqrt(v[k + 1]));
  }
}

generated quantities {
  vector[p] phi;                             // AR coefficients
  real sigma_innovation = sigma * sqrt(v[p + 1]);
  vector[T] mu_fitted = alpha + B * w;
  vector[T] log_lik;                         // for PSIS-LOO

  for (j in 1 : p) {
    phi[j] = a[p, j];
  }

  {
    vector[T] eps = y - mu_fitted;
    for (t in 1 : T) {
      int k = min(t - 1, p);
      real mu = 0;
      for (j in 1 : k) {
        mu += a[k, j] * eps[t - j];
      }
      log_lik[t] = normal_lpdf(eps[t] | mu, sigma * sqrt(v[k + 1]));
    }
  }
}
```

## Fitting with cmdstanr

The driver script is in `fit_rbf_arp.R`. It builds the data list,
compiles once, and then runs the same model under each of CmdStan's
inference algorithms.

The algorithms differ in what they approximate and in what they cost.
Full Hamiltonian Monte Carlo (`$sample()`) targets the exact posterior
and is the reference against which the others should be judged.
`$pathfinder()` runs a quasi-Newton optimisation and draws from a
normal approximation along the trajectory; it is fast and is designed
partly to supply initial values for HMC. `$laplace()` optimises to the
posterior mode and uses the Hessian there, so it is exact for a
Gaussian posterior and degrades smoothly as the posterior departs from
one. `$variational()` (ADVI) fits a factorised or full-rank Gaussian by
stochastic gradient descent, and is the least reliable of the four.
`$optimize()` returns a point estimate only, the penalised maximum
likelihood, with no uncertainty at all.

For this model the posterior should be close to Gaussian in $w$
conditional on the noise parameters, since the mean model is linear.
The departures from normality will be in $\sigma$ and $r$. That
suggests Laplace and Pathfinder will do reasonably well and are worth
using for rapid iteration, with HMC reserved for the final fit. But
that is a hypothesis to check by comparison, not an assumption. The
comparison is the point of running all five.

# What to do next

The single-trial model here is the base case. Three things follow.

Verify the decimation check on a sample of trials and subjects rather
than the one trial used here. One trial is one trial.

Estimate the noise parameters from pooled pre-stimulus data, where
there is no evoked signal, and compare with the whole-epoch estimates.
A material discrepancy would indicate that the whole-epoch noise model
is absorbing signal.

Extend to the multilevel case, where subject-level and trial-level
random functions enter as deviations on the basis weights and the
AR($p$) residual is nested within trial. The `form = ~ idx | trial_id`
grouping, or its Stan equivalent, is essential there: without it the
model will wrap the end of one trial into the start of the next, and
the error is silent.

# References

::: {#refs}
:::
